### 2021-04-26 submission/manuscript.tex
Latest draft of the paper after editing the intro and results

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). However, whether
laxative results in more severe CDIs in mice and how long mice remain
colonized with \emph{C. difficile} after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization. For
FMTs that are delivered via colonoscopy, patients typically undergo
bowel preparation by taking an osmotic laxative prior to the procedure.
Many of the FMT studies to date rationalize the use of laxatives
(20--22) based on a 1996 case study with 2 pediatric patients where the
authors suggested in the discussion that the laxative may help flush
\emph{C. difficile} spores and toxins from the intestine (23).

@@

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 26). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice regained most of
the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 1A). All
treatments rendered mice susceptible to \emph{C. difficile}
colonization. However, PEG-treated mice remained colonized with \emph{C.
difficile} at high levels through thirty days post-infection (Fig. 1C).
In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
within ten days post-infection. Surprisingly, PEG-treated mice were
still susceptible to \emph{C. difficile} infection after 10-days of
recovery from treatment although \emph{C. difficile} was not detectable
in most of the group in the initial five days post-infection (Fig. 1C).
From 9 days post-infection onward, the median \emph{C. difficile}
stabilized for the 10-day recovery group of PEG-treated mice and
remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
laxative treatment alone was sufficient to render mice susceptible to
prolonged \emph{C. difficile} colonization and PEG-treated mice remained
susceptible for up to ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to copraphagy (27), we tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups and primarily housing
only two mice per cage. Importantly, although we conducted a total of 5
separate experiments, the experiment number and it's interaction with
treatment group was not one of the variables that significantly
explained the observed variation in fecal communities (Data Set S1,
Sheet X). Interestingly, none of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-infection
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity index
over time. Although both clindamycin and PEG treatments decreased
diversity, Shannon diversity was lower in the groups of mice that
received PEG treatment compared to those that received clindamycin
through thirty days post-infection (Fig. 2C). We next examined the
bacterial genera that shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-treatment. We found 18 genera that were altered by PEG
treatment (Data Set S1, Sheet X). The majority of the bacterial relative
abundances decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment as \emph{Bacteroides} actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
that differ across treatment groups over multiple timepoints. We found
24 genera were different over multiple timepoints out of the 33 that
were different between treatment groups (Fig. 2E, Data Set S1, Sheet X).
Thus, PEG has a significant impact on the fecal microbiota that was
maintained over time and distinct from clindamycin treatment.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
days post-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). Although we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated that PEG treatment disrupts the mucus
layer and alters the immune response in mice (16), we decided to examine
the impact of PEG treatment on the mucosal microbiota and CDI severity.
To evaluate the mucosal microbiota, we sequenced snips of tissue
collected from the cecum, proximal colon, and distal colon. Similar to
what was observed with the stool samples, alpha diversity was lower in
the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
Although alpha diversity continued to increase over time based on the
communities from PEG-treated mice collected at 20 and 30 days
post-infection (Fig. 3A, Data Set S1, Sheet X). Group, day of the
experiment, and their interactions with other variables (cage,
experiment number, and sample type) explained the majority of the
variation observed in mucosal communities (PERMANOVA combined
R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
Set S1, Sheet X). We saw the greatest difference in the mucosal
microbiota between treatment groups at 6 days post-infection with 16
(insert code instead) genera that were significantly different in all
three of the tissue types we collected (cecum, proximal colon, and
distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days
post-infection only \emph{Bacteroides}, \emph{Clostridiales},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
treatment groups and just in the cecum tissues (Fig. 3C, Fig. 2E, Data
Set S1, Sheet X). Interestingly, one of the genera that was different
between treatment groups at 6 days post-infection was
\emph{Peptostreptococcaceae} (the genera with sequences that match C.
difficile). \emph{Peptostreptococcaceae} was primarily only present in
the 5-day PEG treatment group of mice and decreased in the proximal and
distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a
significant impact on the mucosal microbiota and we detected \emph{C.
difficile} sequences in the cecum, proximal colon, and distal colon
tissue communities.

Next, we examined CDI severity by evaluating cecum and colon
histopathology (28) and found there was no difference in cecum and colon
scores between clindamycin and PEG-treated mice that were challenged
with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the
timepoint typically examined in \emph{C. difficile} 630 challenged mice
(29). We also looked at 6 days post-infection because that was when we
started to see a large difference in C\emph{. difficile} colonization
levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
there was a slight difference in the colon between PEG and
clindamycin-treated mice, there was no difference in the cecum and the
overall score is still relatively low given that the max possible
summary score is 12 (Fig. 3E). Thus, although PEG treatment had a
profound impact on the mucosal microbiota, the impact of \emph{C.
difficile} 630 infection on the cecum and colon was similar between PEG
and clindamycin treated mice.

@@

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption} Next, we
decided to examine how a shorter osmotic laxative perturbation would
impact the microbiome and susceptibility to \emph{C. difficile}. We
administered mice either a 1-day PEG treatment, a 1-day PEG treatment
with a 1-day recovery period, or a clindamycin treatment before
challenging them with \emph{C. difficile 630} (Fig. 3A). In contrast to
the 5-day PEG treated mice, the 1-day PEG treated mice were only
transiently colonized and cleared \emph{C. difficile} by 7 days
post-infection (Fig. 3B). The stool communities of PEG-treated mice were
also only transiently disrupted, with Shannon diversity close to
baseline by 7 days post-infection (Fig. 3C-D). We found 14 bacteria were
impacted by treatment but recovered close to baseline levels within 7
days post-infection including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of treatment matters when considering the impact of
the laxatives, with shorter treatments resulting in a transient loss of
\emph{C. difficile} colonization resistance.

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered} Since 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. We were also interested in exploring whether
PEG might help with engraftment in the context of FMTs. To examine the
impact of PEG treatment on \emph{C. difficile} clearance and FMT
treatment, we treated 4 groups of mice with clindamycin and then
challenged all mice with \emph{C. difficile} before administering the
following treatments: no additional treatment, 1-day PEG immediately
after challenge, and 1-day PEG treatment 3 days after challenge followed
by either administration of an FMT or PBS solution via gavage (Fig. 5A).
In contrast to our hypothesis, all groups of mice that received PEG
actually exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).

\begin{itemize}
\item
  Draft of results

  \begin{itemize}
  \tightlist
  \item
    Prolonged \emph{C. difficile} colonization in post-CDI PEG treated
    mice (Fig. 5B)
  \item
    FMT appears to partially restore alpha diversity (Shannon, but not
    richness Fig. 5C-D)
  \item
    PCoA (necessary?) Could comment on the clustering of Clindamycin and
    PEG-treated mice that received FMT (Fig. 6A)
  \item
    Only 2 genera significantly impacted by FMT treatment, likely not as
    important for clearance (Fig. 6B)
  \item
    Bacteria that consistently differ between groups over time,
    associated with prolonged colonization (Fig. 6C)
  \end{itemize}
\end{itemize}

\textbf{Five-day post-infection community data can predict mice that
will have prolonged \emph{C. difficile} colonization} After identifying
bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG
treatments, we decided to examine the taxa that were influencing
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with input bacterial community data from 5 days
post-infection to predict whether the mice were colonized with \emph{C.
difficile} 10 days post-infection. We chose 5 days post-infection
because that was the earliest timepoint where we would see a treatment
effect in the mice that were given 1-day PEG treatment three days after
\emph{C. difficile} challenge and then administered an FMT or PBS
gavage. The random forest model had the highest performance (AUROC =
0.90, Data Set S1, Sheet X), so we next performed permutation importance
to examine the bacteria that were driving performance. We selected the
top 10 bacteria contributing to our models performance (Fig. 7A) and
examined their relative abundance at 5 days post-infection, the
timepoint used to predict C. difficile colonization status on day 10
(Fig. 7B). Next, we focused on the 5 genera that were \textgreater{} 1
\% relative abundance in either the cleared or colonized mice and
examined the relative abundance dynamics of these bacteria over time. We
found \emph{, }, and \_ tended to have a higher relative abundance and
\textbf{, } had a lower relative abundance in the mice with prolonged
colonization compared to the mice that cleared \emph{C. difficile} (Fig.
7C). Previous work examining the impact of PEG on the murine microbiota
found that PEG treatment resulted in the permanent loss of S24-7, also
known as \emph{Muribaculum intestinale} (16). We decided to check our
\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
Database Project (RDP) database (31), \emph{Porphyromonadaceae} was a
top feature in the random forest model predicting prolonged \emph{C.
difficile} colonization, and had a high relative abundance in the
communities of mice that cleared \emph{C. difficile} within 10 days. We
identified 4 OTUs that had at least (insert minimum percent identity) to
\emph{Muribaculum intestinale} and examined their abundance in mice that
either cleared or were still colonized with \emph{C. difficile} at 10
days post-infection (Fig. S4).

@@

\textbf{Figure 1. 5-day PEG treatment prolongs susceptibility and mice
become persistently colonized with \emph{C. difficile}.} A. Setup of the
experimental timeline for experiments with 5-day PEG treated mice.
Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
15\% PEG 3350 was administered in the drinking water for five days. B.
Weight change from baseline weight in groups after treatment with PEG
and/or clindamycin, followed by \emph{C. difficile} challenge. C.
\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
per timepoint) via serial dilutions. The black line represents the limit
of detection for the first serial dilution. CFU quantification data was
not available for each mouse due to stool sampling difficulties
(particularly the day the mice came off of the PEG treatment) or early
deaths. For B-C, lines represent the median for each treatment group and
circles represent samples from individual mice. Asterisks indicate
timepoints where the weight change or CFU/g was significantly different
between groups by the Kruskal-Wallis test with Benjamini-Hochberg
correction for testing multiple timepoints. The data presented are from
a total of 5 separate experiments. \newpage

@@

\includegraphics{figure_4.pdf} \textbf{Figure 4. 1-day PEG treatment
renders mice susceptible to transient \emph{C. difficile} colonization.}
A. Setup of the experimental timeline for the 1-day PEG treated subset
of mice. B. CFU/gram stool measured over time (N = 12-18 mice per
timepoint) via several dilutions. The black dotted line represents the
limit of detection for the first serial dilution. Asterisks indicate
timepoints where the CFU/gram was significantly different between groups
using the Kruskall-Wallis test with a Benjamini-Hochberg correction for
multiple timepoints. C. Principle Coordinate Analysis plot of the groups
over time with the alpha representating the same time scale as in panel
D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
0.19). D. Shannon diversity Index of the groups over time. Only days
with samples from all groups are shown. E. Line plots of relative
percent abundance of selected genera over time. Only days with samples
from all groups shown. The gray line represents the limit of detection.

>>>>

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). However, whether
laxative administration results in more severe CDIs in mice and how long
mice remain colonized with \emph{C. difficile} after challenge is
unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization
resistance. For FMTs that are delivered via colonoscopy, patients
typically undergo bowel preparation by taking an osmotic laxative prior
to the procedure. Many of the FMT studies to date rationalize the use of
laxatives (20--22) based on a 1996 case study with 2 pediatric patients
where the authors suggested in the discussion that the laxative may help
flush \emph{C. difficile} spores and toxins from the intestine (23).

@@

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 26). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice recovered most
of the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 1A). All
treatments rendered mice susceptible to \emph{C. difficile}
colonization. However, PEG-treated mice remained colonized with \emph{C.
difficile} at a high level through thirty days post-infection (Fig. 1C).
In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
within ten days post-infection. Surprisingly, PEG-treated mice were
still susceptible to \emph{C. difficile} infection after 10-days of
recovery from treatment, although \emph{C. difficile} was not detectable
in most of the group in the initial five days post-infection (Fig. 1C).
From 8 days post-infection onward, the median \emph{C. difficile} CFU
stabilized for the 10-day recovery group of PEG-treated mice and
remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
laxative treatment alone was sufficient to render mice susceptible to
prolonged \emph{C. difficile} colonization and PEG-treated mice remained
susceptible up to ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to copraphagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups and primarily housing
only two mice per cage. Importantly, although we conducted a total of 5
separate experiments, the experiment number and it's interaction with
treatment group was not one of the variables that significantly
explained the observed variation in fecal communities (Data Set S1,
Sheet X). Interestingly, none of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-infection
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity over
time. Although both clindamycin and PEG treatments decreased diversity,
Shannon diversity was lower in the groups of mice that received PEG
treatment compared to those that received clindamycin through thirty
days post-infection (Fig. 2C). We next examined the bacterial genera
that shifted after PEG treatment by comparing the baseline samples of
mice treated with only PEG to samples from the same mice one day
post-treatment. We found 18 genera that were altered by PEG treatment
(Data Set S1, Sheet X). The majority of the bacterial relative
abundances decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment, as \emph{Bacteroides} actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
that differed across treatment groups over multiple time points. We
found 24 genera were different over multiple time points out of the 33
genera that were different between treatment groups (Fig. 2E, Data Set
S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota
that was maintained over time and was distinct from clindamycin
treatment.

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
days post-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). While we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although
alpha diversity continued to increase over time based on the communities
from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
3A, Data Set S1, Sheet X). Group, time point, and their interactions
with other variables (cage, experiment number, and sample type)
explained the majority of the variation observed in mucosal communities
(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
0.05, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in
the mucosal microbiota between treatment groups at 6 days post-infection
with 16 genera that were significantly different in all three of the
tissue types we collected (cecum, proximal colon, and distal colon; Fig.
S3A, Data Set S1, Sheet X). However, at 30 days post-infection only
\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
\emph{Ruminococcaceae} were different between treatment groups and only
in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X).
Interestingly, one of the bacteria that was different between treatment
groups at 6 days post-infection was \emph{Peptostreptococcaceae} (the
genera with an OTU sequence that matches \emph{C. difficile}).
\emph{Peptostreptococcaceae} was primarily only present in the 5-day PEG
treatment group of mice and decreased in the proximal and distal colon
tissues over time (Fig. S2B). Thus, PEG treatment had a significant
impact on the mucosal microbiota and we detected \emph{C. difficile}
sequences in the cecum, proximal colon, and distal colon tissue
communities.

Next, we examined CDI severity by evaluating cecum and colon
histopathology (28) and found there was no difference in cecum and colon
scores between clindamycin and PEG-treated mice that were challenged
with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the time
point typically examined in \emph{C. difficile} 630 challenged mice
(29). We also looked at 6 days post-infection because that was when we
started to see a large difference in \emph{C. difficile} colonization
levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
there was a slight difference in the colon between PEG and
clindamycin-treated mice, there was no difference in the cecum and the
overall score is still relatively low given that the max possible
summary score is twelve (Fig. 3E). Therefore, although PEG treatment had
a disruptive effect on the mucosal microbiota, the impact of \emph{C.
difficile} 630 infection on the cecum and colon was similar between PEG
and clindamycin treated mice.

@@

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption} Next, we
decided to examine how a shorter osmotic laxative perturbation would
impact the microbiome and susceptibility to \emph{C. difficile}. We
administered either a 1-day PEG treatment, a 1-day PEG treatment with a
1-day recovery period, or clindamycin to mice before challenging them
with \emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
mice, the 1-day PEG treated mice were only transiently colonized and
cleared \emph{C. difficile} by 7 days post-infection (Fig. 3B). The
stool communities of PEG-treated mice were also only transiently
disrupted, with Shannon diversity recovering close to baseline by 7 days
post-infection (Fig. 3C-D). We found 14 bacteria were impacted by
treatment, but recovered close to baseline levels by 7 days
post-infection including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of treatment matters when considering the impact of
the laxatives, with shorter treatments resulting in a transient loss of
\emph{C. difficile} colonization resistance.

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered} Since 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. We were also interested in exploring whether
PEG might help with engraftment in the context of FMTs. To examine the
impact of PEG treatment on \emph{C. difficile} clearance and FMT
treatment, we treated 4 groups of mice with clindamycin and then
challenged all mice with \emph{C. difficile} before administering the
following treatments: no additional treatment, 1-day PEG immediately
after challenge, and 1-day PEG treatment 3 days after challenge followed
by either administration of an FMT or PBS solution by oral gavage (Fig.
5A). Contrary to our hypothesis, all groups of mice that received PEG
exhibited prolonged \emph{C. difficile} colonization (Fig. 5B). Next we
examined how post-CDI PEG treatment impacted the microbiota. Alpha
diversity was lower in the PEG-treated mice with the exception of the
PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT
appeared to partially restore Shannon diversity but not richness (Fig.
5C-D). Similarly, we saw some overlap between the communities of mice
that received FMT and the mice treated with only clindamycin after 5
days post-infection (Fig. 6A). The increase in Shannon diversity
suggests that the FMT did have an impact on the microbiota, despite
seeing prolonged \emph{C. difficile} colonization in the FMT treated
mice. However, only \emph{Bacteroidales} and \emph{Porphyromonadaceae}
consistently differed between the mice received either an FMT or PBS
gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
Overall, we found there were 24 bacteria that differed over multiple
time points between treatment groups including increased
\emph{Akkermansia} and decreased \emph{Ruminococcaceae},
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Oscillibacter}
in mice that received PEG after \emph{C. difficile} challenge (Fig. 6C).
In sum, administering PEG actually prolonged \emph{C. difficile}
colonization, including in mice that received an FMT, which only
restored a couple of bacterial genera.

\textbf{Five-day post-infection community data can predict which mice
that will have prolonged \emph{C. difficile} colonization} After
identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day
PEG treatments, we decided to examine the bacteria that influenced
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with input bacterial community data from 5 days
post-infection to predict whether the mice were still colonized with
\emph{C. difficile} 10 days post-infection. We chose 5 days
post-infection because that was the earliest time point where we would
see a treatment effect in the mice that were given 1-day PEG treatment
three days after \emph{C. difficile} challenge and then administered an
FMT or PBS gavage. The random forest model had the highest performance
(AUROC = 0.90, Data Set S1, Sheet X), so we next performed permutation
importance to examine the bacteria that were the top contributors to the
random forest model predicting prolonged \emph{C. difficile
}colonization. We selected the top 10 bacteria contributing to our
models performance (Fig. 7A) and examined their relative abundance at 5
days post-infection, the time point used to predict \emph{C. difficile}
colonization status on day 10 (Fig. 7B). Next, we focused on the 5
genera that had a greater than 1 \% relative abundance in either the
cleared or colonized mice and examined how the bacteria changed over
time. We found \emph{Enterobacteriaceae} and \emph{Bacteroides} tended
to have a higher relative abundance, \emph{Akkermansia} was initially
decreased and then increased, and \emph{Porphyromonadaceae} and
\emph{Lachnospiraceae} had a lower relative abundance in the mice with
prolonged colonization compared to the mice that cleared \emph{C.
difficile} (Fig. 7C). Together these results suggest a combination of
low and high abundance bacterial genera influence the prolonged
colonization observed in 5-day PEG and post-CDI PEG treated mice.

Previous work examining the impact of PEG on the murine microbiota found
that PEG treatment resulted in the permanent loss of S24-7, also known
as \emph{Muribaculum intestinale} (16). We decided to check our
\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
classified as \emph{Porphyromonadaceae} by the Ribosomal Database
Project (RDP) database (31) and \emph{Porphyromonadaceae} was a top
feature in the random forest model predicting prolonged \emph{C.
difficile} colonization. We identified 4 OTUs that had at least 92\%
identity to \emph{Muribaculum intestinale} and examined their abundance
in mice that either cleared or were still colonized with \emph{C.
difficile} at 10 days post-infection. While all of the OTUs, were
decreased by PEG and clindamycin treatment, there was some recovery in
the mice that cleared (Fig. S4A). We also examined other
\emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs since these
were the 2 genera that were important to our classification model and
contained multiple OTUs that were different at 5 days post-infection
between mice that either cleared or remained colonized with \emph{C.
difficile} by 10 days post-infection (Data Set S1, sheet X). While
individual \emph{Porphyromonadaceae} and \emph{Lachnospiraceae} OTUs
tended to be more abundant in the mice that clear \emph{C. difficile}
relative to the mice that exhibit prolonged colonization (Fig. S4B-C),
there is no single OTU that fits the pattern we observed at the genus
level (Fig. 7C), suggesting multiple \emph{Porphyromonadaceae} and
\emph{Lachnospiraceae} OTUs influenced \emph{C. difficile} clearance.
Overall, our results suggest that specific bacterial community
differences explain the prolonged \emph{C. difficile} colonization we
observed in 5-day PEG and post-CDI 1-day PEG treated mice.

@@

\textbf{Figure 1. 5-day PEG treatment prolongs susceptibility and mice
become persistently colonized with \emph{C. difficile}.} A. Setup of the
experimental timeline for experiments with 5-day PEG treated mice.
Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
15\% PEG 3350 was administered in the drinking water for five days. B.
Weight change from baseline weight in groups after treatment with PEG
and/or clindamycin, followed by \emph{C. difficile} challenge. C.
\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
per time point) via serial dilutions. The black line represents the
limit of detection for the first serial dilution. CFU quantification
data was not available for each mouse due to stool sampling difficulties
(particularly the day the mice came off of the PEG treatment) or early
deaths. For B-C, lines represent the median for each treatment group and
circles represent samples from individual mice. Asterisks indicate time
points where the weight change or CFU/g was significantly different
between groups by the Kruskal-Wallis test with Benjamini-Hochberg
correction for testing multiple time points. The data presented are from
a total of 5 separate experiments. \newpage

@@

\includegraphics{figure_4.pdf} \textbf{Figure 4. 1-day PEG treatment
renders mice susceptible to transient \emph{C. difficile} colonization.}
A. Setup of the experimental timeline for the 1-day PEG treated subset
of mice. B. CFU/gram stool measured over time (N = 12-18 mice per time
point) via several dilutions. The black dotted line represents the limit
of detection for the first serial dilution. Asterisks indicate time
points where the CFU/gram was significantly different between groups
using the Kruskall-Wallis test with a Benjamini-Hochberg correction for
multiple time points. C. Principle Coordinate Analysis plot of the
groups over time with the alpha representating the same time scale as in
panel D (day: R\textsuperscript{2} = 0.43; group: R\textsuperscript{2} =
0.19). D. Shannon diversity Index of the groups over time. Only days
with samples from all groups are shown. E. Line plots of relative
percent abundance of selected genera over time. Only days with samples
from all groups shown. The gray line represents the limit of detection.
